---
title: "Project 1"
author: "Garrett Storey"
date: "2022-09-16"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(plotly)
library(ggplot2)
library(knitr)
library(rmarkdown)
data <- read.csv("hourly_spec.csv")
```

## Investigating Particulate Matter Concentrations Using EPA Hourly Datasets

```{r include=FALSE}
head(data)
```
### Description of Variables
* State Code
  + FIPS state-level geographic identifier code (integer)
* Country Code
  + FIPS country-level geographic identifier code (integer)
* Site Number
  + EPA-designated site number to identify monitoring stations (integer)
* Parameter Code
  + EPA-designated code corresponding to a unique analyse (integer)
* POC
  + Parameter Occurance Code used to identify if there is more than one device measuring the same pollutant at a site (integer)
* Latitude
  + Distance north or south of equator, expressed as a (double)
* Longitude
  + Distance east or west of the meridian, expressed as a (double)
* Datum
  + Reference frame coordinate system (chr)
* Parameter Name (IMPORTANT)
  + Name of the analyte (chr)
* Date Local
  + Date in the form of "year-month-day" (chr)
* Time Local
  + Time at the monitoring site in military format "00:00" (chr)
* Date GMT
  + Date in Greenwich Mean Time (chr)
* Time GMT
  + Time in Greenwich Mean Time (chr)
* Sample Measurement (IMPORTANT)
  + Sample measurement magnitude (double)
* Units of Measure (IMPORTANT)
  + The units the sample measurement was made in (chr)
* MDL
  + Method detection limit in the units of measure (double)
* 